Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-12-07 document/submission/ms.tex
put in caveats about effective temperature and photospheric temperature, change some language regarding the indexing

We took two approaches to characterizing the effective temperature response from a spotted-star spectrum.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}.  } to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models observed spectra with synthetic spectra from pre-computed model grids.  The intra-grid-point spectra are ``emulated'' in a process similar but superior to interpolation, since it seemlessly quantifies the uncertainty attributable to the coarsely sampled stellar intrinsic parameters (see Appendix of \iancze).  The forward model includes calibration parameters, line spread functions, and a Gaussian process noise model to account for correlations in the residual spectrum.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra, which span a wide range of wavelengths at high spectral resolution with a sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to include a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added by the starspot model.

@@

We performed spectral fitting on an ESPaDOnS spectrum acquired on 2014 January 11.  The spectrum was split into subsets of $N=35$ chunks, corresponding approximately to spectral order boundaries.  Fitting was performed separately on each of 35 spectral orders following the Metropolis-Hasting MCMC sampling procedure described in \iancze.  The spectral emulator was trained on stellar parameters in the range $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and $\teff \in [3500, 4200]$.

The Metropolis-Hastings step sizes were tuned with several iterations of burn-in procedure and the final chains were visually checked for convergence, dropping 9 orders that failed to converge for numerical reasons.  We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall the 26 sets of point estimates for stellar parameters show relatively good agreement, with exceptions.  The best-fitting spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each spectral order.

@@

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in optical.  The most conspicuous trend is in the derived effective temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The effective temperature peaks at values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived effective temperatures of $\sim3300$ K, or 700 K cooler than estimated from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  
Sources with such discrepancies have been seen previously, for example in Figures 4 and 5 in \citet{bouvier92}.
  These discrepancies are circumstantial evidence for the detection of spectral features attributable to starspots.

@@

\begin{figure*}
 \centering
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.5cm, clip=true, width=0.65\textwidth]{lkca4_hrdiag_paper2}
 \caption{Locations for LkCa 4 on an HR diagram, compared with models of pre-main sequence evolution calculated by \citet{baraffe15} with isochrones (black lines) and evolution models of a single mass (dashed blue lines) as marked.  The measured effective temperature and luminosity from this paper, based on the two-component fit and a median $V$-band magnitude, corresponds to the black asterisk.  The yellow shaded region corresponds to the range of apparent effective temperatures that are would be measured as the hot component rotates into and out of our view.  The blue circle corresponds to the measurement at blue-optical wavelengths by \citet{donati14}, the purple square corresponds to the measurement from low-resolution optical spectra, biased to TiO band strengths, by \citet[][biased to]{herczeg14}, while the red diamond corresponds to what we would measure from the K-band spectrum and 2MASS $J$-band magnitude.}
 \label{fig:hrdiag}
\end{figure*}

@@

The effective temperature $T_{\rm eff}$ of these two components may be calculated as
\begin{equation}
T_{\rm eff}=[T_{\rm amb}^4(1-f_{\mathrm{cool}})+T_{spot}^4f_{\mathrm{cool}}]^{0.25}.
\end{equation}
For the median fill factor of 80\%, the observed effective temperature would be 3180 K, with a range of 3070--3280 for 86\% and 74\% spot fill factors.  If the cool spot is 3000 K, at the limit of the acceptable temperature range, then these values would increase by 150 K.  The apparent effective temperature in any single epoch depends on the relative fractions of the hot and cool spot.  When accounting for the visible contributions of both hot and cool components, this wide range of possible observed effective temperatures at any single epoch underscores the difficulty in placing young stars on HR diagrams.

@@

When the photosphere is assumed to be a single temperature, the effective temperature depends on the wavelength where the measurement is made.  For LkCa 4, the $K$-band temperature  is closest to the effective temperature of the entire photosphere.  These discrepancies may also explain the large temperature, systematic offsets identified when comparing temperatures from $H$-band APOGEE \footnote{The APOGEE spectra have a long wavelength cutoff at 1.70 $\mu$m, just short of an IGRINS ``sweet spot'' from $17075-17990 \;\AA$ (orders 100, 101, and 104), which have conspicuous hot and cool features, is far-enough away from the $K-$band where problems like magnetic field-sensitive lines, enhanced telluric absorption, and veiling by disks may complicate the spectra.} spectra to those obtained from other (usually optical) methods \citep{cottaar14}. 
On the other hand, GAIA-ESO spectra \citep{frasca15} cover only a narrow wavelength around H$\alpha$ and would yield an effective temperature of $\sim 4000$ K.   Spectral types and effective temperatures measured from TiO depths \citep{herczeg14} are lower than those from blue spectra but still overestimate the effective temperature.  Temperature uncertainties for young K and M dwarfs are larger than any formal uncertainties in individual measurements.

>>>>

We took two approaches to characterizing the photospheric temperature from a spotted-star spectrum.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}.  } to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models observed spectra with synthetic spectra from pre-computed model grids.  The intra-grid-point spectra are ``emulated'' in a process similar but superior to interpolation, since it seemlessly quantifies the uncertainty attributable to the coarsely sampled stellar intrinsic parameters (see Appendix of \iancze).  The forward model includes calibration parameters, line spread functions, and a Gaussian process noise model to account for correlations in the residual spectrum.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra, which span a wide range of wavelengths at high spectral resolution with a sampling of 100 K in stellar photospheric temperature\footnote{Single component \PHOENIX\ photospheres are indexed by the label ``effective temperature'', or $\teff$, which presupposes that the emergent spectrum is comprised of only a single photospheric component.  In this work we avoid labeling the photospheric components by $\teff$, and instead refer to them as simply ``photospheric temperatures''.} in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to include a two photospheric components.  Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added by the two component model.

@@

We performed spectral fitting on an ESPaDOnS spectrum acquired on 2014 January 11.  The spectrum was split into subsets of $N=35$ chunks, corresponding approximately to spectral order boundaries.  Fitting was performed separately on each of 35 spectral orders following the Metropolis-Hasting MCMC sampling procedure described in \iancze.  The spectral emulator was trained on stellar parameters in the range $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and photospheric temperatures $\in [3500, 4200]$.

The Metropolis-Hastings step sizes were tuned with several iterations of burn-in procedure and the final chains were visually checked for convergence, dropping 9 orders that failed to converge for numerical reasons.  We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall the 26 sets of point estimates for stellar parameters show relatively good agreement, with exceptions.  The best-fitting spectral orders show photospheric temperature point estimates in the range $4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each spectral order.

@@

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in optical.  The most conspicuous trend is in the derived photospheric temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The photospheric temperature peaks at values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived photospheric temperatures of $\sim3300$ K, or 700 K cooler than estimated from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  
Sources with such discrepancies have been seen previously, for example in Figures 4 and 5 in \citet{bouvier92}.  These discrepancies are circumstantial evidence for the detection of spectral features attributable to a second photospheric component.

@@

\begin{figure*}
 \centering
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.5cm, clip=true, width=0.65\textwidth]{lkca4_hrdiag_paper2}
 \caption{Locations for LkCa 4 on an HR diagram, compared with models of pre-main sequence evolution calculated by \citet{baraffe15} with isochrones (black lines) and evolution models of a single mass (dashed blue lines) as marked.  The measured effective temperature and luminosity from this paper, based on the two-component fit and a median $V$-band magnitude, corresponds to the black asterisk.  The yellow shaded region corresponds to the range of apparent effective temperatures that would be measured as the hot component rotates into and out of our view.  The blue circle corresponds to the measurement at blue-optical wavelengths by \citet{donati14}, the purple square corresponds to the measurement from low-resolution optical spectra, biased to TiO band strengths, by \citet[][biased to]{herczeg14}, while the red diamond corresponds to what we would measure from the K-band spectrum and 2MASS $J$-band magnitude.}
 \label{fig:hrdiag}
\end{figure*}

@@

The effective temperature $T_{\rm eff}$ of these two components may be calculated as
\begin{equation}
T_{\rm eff}=[T_{\rm hot}^4(1-f_{\mathrm{cool}})+T_{\mathrm{cool}}^4 f_{\mathrm{cool}}]^{0.25}.
\end{equation}
For the median fill factor of 80\%, the observed effective temperature would be 3180 K, with a range of 3070--3280 for 86\% and 74\% spot fill factors.  If the cool spot is 3000 K, at the limit of the acceptable temperature range, then these values would increase by 150 K.  The apparent effective temperature in any single epoch depends on the relative fractions of the hot and cool spot.  When accounting for the visible contributions of both hot and cool components, this wide range of possible observed effective temperatures at any single epoch underscores the difficulty in placing young stars on HR diagrams.

@@

When the photosphere is assumed to be a single temperature, the effective temperature depends on the wavelength where the measurement is made.  For LkCa 4, the $K$-band temperature  is closest to the effective temperature of the entire photosphere.  These discrepancies may also explain the large temperature, systematic offsets identified when comparing temperatures from $H$-band APOGEE spectra to those obtained from other (usually optical) methods \citep{cottaar14}.  Two component photopheres would also explain deriving different spectral types in the optical and infrared for the same source.
On the other hand, GAIA-ESO spectra \citep{frasca15} cover only a narrow wavelength around H$\alpha$ and would yield an effective temperature of $\sim 4000$ K.   Spectral types and effective temperatures measured from TiO depths \citep{herczeg14} are lower than those from blue spectra but still overestimate the effective temperature.  Temperature uncertainties for young K and M dwarfs are larger than any formal uncertainties in individual measurements.
